\speculake: Hiding Malware in Speculative Execution

### 2018-08-02 paper/payloads.tex
first pass on turing and decrypt

As a first step, we observe that the speculative primitive can be used to trivially
implement a finite state machine: any logic can be done in the speculative
world, while updates to the state are communicated to the real world where they
are stored. On the next run of the speculative instructions, they read from the
real world state (and other inputs), compute any state transitions, and
communicate the next state. In this mode, the state is maintained by the real world,
while updates are controlled by code executed speculatively.

Next, we observe that the instructions to be executed speculatively can be
obscured further by encrypting them. This encryption allows for hidden
computation, instructions are stored in the real world encrypted, they are then
decrypted in the speculative world and passed to the real world to be executed.
This method makes it appear as though the program is generating its own code.

@@

\subsection{Turing Machine}
\label{subsec:turing}
To demonstrate that arbitrary computation can be done in the speculative world
we implement a Turing machine, configured to run the 5-state Busy Beaver 
function~\cite{chaitin1987computing,herken1992universal}.
The Busy Beaver is a configuration which writes the most
1s on the tape before halting with a limited number of states.

The computation of the Turing Machine is done speculatively, meaning that while
the real world keeps track of the state, tape, and symbols of the turing machine
all logic used to update the above is done in the speculative world. Figure~\ref{fig:general_model} demonstrates the execution flow of a sample \speculake malware.

@@

While a Turing machine demonstrates that arbitrary speculative computation
is possible, hiding malware this way has several drawbacks. First, an
analyst might be able to observe that something is driving a Turing
machine by noticing the real world state updating. With enough symbols, the
analyst could reverse engineer the state machine even without being able to see
the internal logic executing speculatively. Second, Turing machines are a
poor choice for practical computation, as they are inefficient and have no direct
way to perform system calls or otherwise interface externally.

We explore a more practical application of using \speculake to perform
decryption speculatively. To hide keys from the analyst, the key and decryption
code only occur in the speculative world, while committed state only learns the
resulting plaintext. Furthermore, the speculative code can
derive the decryption key based entirely off the trigger program, making it
impossible for an analyst that only has the payload program to
determine the key or decrypt the embedded ciphertext.

Any observer watching the execution of this program would never see any
indication of decryption, due to the exploitation of speculative execution. Any
observation of the executed instructions would not include any decryption
instructions, or loading of keys as they are only done speculatively, and thus
not shown in any debugging or real world traces of program execution. 
%Additionally the observer would be unable to find the executed instructions in
%the ELF file as they are only stored encrypted.

>>>>

As a first step, we observe that the speculative primitive can be used to trivially
implement a finite state machine: logic can be done in the speculative
world, while updates to the state are communicated to the real world where they
are stored. On the next run of the speculative instructions, the state is read from the
real world state (along with any inputs), state transitions are computed and communicated back.
In this mode, the state is maintained by the real world,
while updates are controlled by code executed speculatively.

We further observe we are not limited to finite state machines, but can support
any model of computation where \emph{updates} to any state are finite (i.e. can fit
within the bandwidth constraints of the speculative primitive). This
encompases Turing machines~\cite{turing1937computable} as well as certain random access machines, which we
investigate next.
Figure~\ref{fig:general_model} demonstrates the execution flow of a sample \speculake malware.

%Next, we observe that the instructions to be executed speculatively can be
%obscured further by encrypting them. This encryption allows for hidden
%computation, instructions are stored in the real world encrypted, they are then
%decrypted in the speculative world and passed to the real world to be executed.
%This method makes it appear as though the program is generating its own code.

@@

\subsection{Turing Machine}
\label{subsec:turing}
To demonstrate that arbitrary computation can be performed cooperatively between
the speculative world and real world,
we implement a Turing machine, and configure it to run a 5-state Busy Beaver 
function~\cite{turing1937computable,chaitin1987computing,herken1992universal}. This configuration allows
us to run a large number of steps with very minimal logic.
%The Busy Beaver is a configuration which writes the most
%1s on the tape before halting with a limited number of states.

Updates to this Turing machine are computed speculatively, while the real
world keeps track of the state and full tape of the machine. Thus, the logic of
the machine is entirely contained in the speculative world, while the state may
be externally visible (e.g. to a dynamic debugger). We note that the machine
only operates when the trigger executes, making it difficult for an analyst with
only access to the Turing machine to determine exactly what the machine will do
from its initial state.

However, this toy example is meant only as an illustrative example of arbitrary
computation, notas a robust means of obfuscation. Indeed, even the initial state
of a Turing machine alone may reveal a significant amount of information.
Furthermore, the analyst may attempt to locate potential speculative entry
points, even without the help of the trigger program. We describe ways to address
both of these next.

@@

While a Turing machine demonstrates that arbitrary speculative computation
is possible, hiding malware this way has several drawbacks. First, Turing
machines are a poor choice for practical computation, as they are inefficient
and have no direct way to interface with the rest of the system (e.g. via system
calls). Second, as mentioned, they leave a great deal of information available
to the analyst, including the initial tape state, and a potentially small
(enumerable) number of possible speculative entry points.

We explore a more practical application of using \speculake to perform
\emph{decryption} speculatively. To hide keys from the analyst, the key and decryption
code only occur in the speculative world, while the initial payload program
contains only the ciphertext. While partial plaintext will be available in the
real world during execution, we emphasize that this only occurs when the trigger
runs. Before this, the state of the program reveals only the ciphertext that
will be decrypted. While the speculative entry point enumeration attack could be
used to reveal the keys used to decrypt this ciphertext, we describe a way to
derive the decryption key entirely from the trigger program. Thus, an analyst
that only has access to the payload program will be unable to learn the key or
decrypt the embedded ciphertext.

We also note that even when the trigger runs, decryption does not occur outside
the speculative context, meaning that any traditional traps or debugging
breakpoints placed on decryption instructions or routines will not occur, even
as they are used speculatively. These instructions could even be obfuscated themselves by
placing them in other misaligned instructions, and choosing a speculative entry
point that jumps to the middle of other instructions.
